\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add conf bands

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: 
  rticles::rjournal_article:
    number_sections: true
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area. The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

```{r, eval = TRUE, echo=eval_cs1}
summary(fit) 
```

@@

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
# Recompute with more time points
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     ylab = "Cumulative incidence (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
```

@@

In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available. For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve. The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals. Furthermore, we would like to include more functions to compute calibration and discrimination statistics (e.g. AUC) for our models. Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model. Their approach could potentially be included in \pkg{casebase}. Finally, we want to provide more flexibility in how the case-base sampling is performed. This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}. In this way, users could specify their own sampling mechanism. For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].

@@

We would like to thank Yi Yang for helpful discussions on penalized regression models and gradient boosting. \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area. The case-base sampling approach described above can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

```{r, eval = eval_cs1, echo = TRUE}
summary(fit) 
```

@@

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
# Recompute with more time points
risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     ylab = "Cumulative incidence (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
```

Moreover, we can also compute confidence intervals for the CIF using the method \code{confint.absRiskCB}.

```{r eval = eval_cs1, echo = -1}
new_time <- c(0, 1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the CIF estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

# plot absrisk and 95% CI
new_time <- seq(0,14,0.1)
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

x.poly <- c(new_time[i.forw], new_time[i.backw])
range_y <- 100 * range(c(tt$conf.high, tt$conf.low)) * c(0.99, 1.01)

par(mfrow = c(1, 2))
# 1. Control
tt_sub <- subset(tt, cov_prof == "Control group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        ylab = "Cumulative incidence (%)",
        xlab = "Years since randomization",
        main = "Control group"
    )
)

graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)

# Add KM
km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
                  data = ERSPC,
                  subset = ScrArm == "Control group",
                  conf.type = "log-log")

lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
      lwd = 1, lty = 1, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
      lwd = 2, lty = 3, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
      lwd = 2, lty = 3, col = myCols[2])

do.call("lines", 
        list(
            x = new_time[i.forw],
            y = 100 * tt_sub$estimate[i.forw],
            lwd = 2,
            lty = 1,
            col = myCols[1]
        )
)

# Add legend
legend("topleft", 
       legend = c("Case-base", "Kaplan-Meier"), 
       col = myCols,
       lty = c(1, 1), 
       lwd = c(2, 1),
       cex = 1.1,
       bg = "gray90")

# 2. Screening
tt_sub <- subset(tt, cov_prof == "Screening group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        # ylab = "Cumulative incidence (%)",
        ylab = "",
        xlab = "Years since randomization",
        main = "Screening group"
    )
)

graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)

# Add KM
km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
                  data = ERSPC,
                  subset = ScrArm == "Screening group",
                  conf.type = "log-log")

lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
      lwd = 1, lty = 1, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
      lwd = 2, lty = 3, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
      lwd = 2, lty = 3, col = myCols[2])

do.call("lines", 
        list(
            x = new_time[i.forw],
            y = 100 * tt_sub$estimate[i.forw],
            lwd = 2,
            lty = 1,
            col = myCols[1]
        )
)
```

@@

In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available. For example, we would like to include more functions to compute calibration and discrimination statistics (e.g. AUC) for our models. Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model. Their approach could potentially be included in \pkg{casebase}. Finally, we want to provide more flexibility in how the case-base sampling is performed. This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}. In this way, users could specify their own sampling mechanism. For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].

@@

We would like to thank the anonymous reviewers for their insightful comments and criticisms. We would also like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add conf bands

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{table}

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen. Cox), and our approach using penalized case-base sampling (Pen. CB). Only the covariates that were selected by both penalized approaches are shown. The shrinkage of the coefficient estimates for Pen. Cox and Pen. CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen. Cox), penalized case-base sampling estimation (Pen. CB), and Kaplan-Meier (K-M). (A)  Probability of death as a function of follow-up time. (B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
\end{figure}
\end{Schunk}

@@

In future versions of the package, we also want to increase the
complement of diagnostic and inferential tools that are currently
available. For example, we would like to include the ability to compute
confidence intervals for the cumulative incidence curve. The delta
method or parametric bootstrap are two different strategies we can use
to construct approximate confidence intervals. Furthermore, we would
like to include more functions to compute calibration and discrimination
statistics (e.g.~AUC) for our models. Saarela and Arjas
\citeyearpar{saarela2015non} also describe how to obtain a posterior
distribution for the AUC from their model. Their approach could
potentially be included in \pkg{casebase}. Finally, we want to provide
more flexibility in how the case-base sampling is performed. This could
be achieved by adding a \code{hazard} argument to the function
\code{sampleCaseBase}. In this way, users could specify their own
sampling mechanism. For example, they could provide a hazard that gives
sampling probabilities that are proportional to the cardiovascular
disease event rate given by the Framingham score \citep{saarela2015non}.

>>>>

Moreover, we can also compute confidence intervals for the CIF using the
method \code{confint.absRiskCB}.

\begin{Schunk}
\begin{Sinput}
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high        cov_prof
#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
\end{Soutput}
\end{Schunk}

These point-wise confidence intervals are constructed using parametric
bootstrap; the validity of this approach relies on the asymptotic
normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
95\% confidence bands around the CIF estimates.

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
\end{figure}
\end{Schunk}

@@

@@

@@

@@

@@

@@

In future versions of the package, we also want to increase the
complement of diagnostic and inferential tools that are currently
available. For example, we would like to include more functions to
compute calibration and discrimination statistics (e.g.~AUC) for our
models. Saarela and Arjas \citeyearpar{saarela2015non} also describe how
to obtain a posterior distribution for the AUC from their model. Their
approach could potentially be included in \pkg{casebase}. Finally, we
want to provide more flexibility in how the case-base sampling is
performed. This could be achieved by adding a \code{hazard} argument to
the function \code{sampleCaseBase}. In this way, users could specify
their own sampling mechanism. For example, they could provide a hazard
that gives sampling probabilities that are proportional to the
cardiovascular disease event rate given by the Framingham score
\citep{saarela2015non}.
